Trauma treatment delay

### 2023-12-14 manuscriptlongversion.Rmd
Comment on manuscript

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(knitr)

```

@@

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, e.g. delayed treatment. *Aim:* This
retrospective cohort study aims to identify patient level factors that
are linked with receiving delayed treatment among trauma patients.
*Material and Methods:* This study uses the database of Karolinska
university hospital, the equivalent of a level 1 trauma center, to run a
logistical regression linking the outcome, delayed treatment, to
variables, observed patient level factors. Patients treated for trauma
at the hospital between 2012 and 2022 were included, with ISS\>9 and/or
trauma code activation. Exclusion criterias were not being screened for
ofi and missing values in the used variables, age, gender, ISS, SBP, RR,
GCS and number of injuries. *Results:* Higher ISS and RR was linked to
higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not when adjusted for. The
remaining variables showed no correlation. *Conclusion:* Our study
suggests that it is mostly more severely injured and less awake patients
that receive delayed treatment.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response [@gerdin_risk_2015], is and has long been
a major cause of death around the world [@WHO_2021]. Within modern
hospital care, the trauma system stands as an important part in
healthcare and is crucial to lowering mortality and morbidity in injured
patients through many means, such as hospital care, patient follow up
and prevention programs [@Choi_2021].

## Trauma Statistics

Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
groups, as blunt trauma patients tend to be more injured on arrival to
the hospital and as such, require more resources and are hospitalized
for a longer periods [@FITCH2019].

@@

## Trauma systems

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012] . During the
last three decades, the introduction of trauma systems has contributed
to lowering the incidence of preventable death . This is attributed to
the improvements in care for acute brain injuries and bleeding control.
The incidence of late death because of sepsis and multiple organ failure
has also been lowered, possibly a result of better and earlier
resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
the hospital, but their work outside of the hospital is just as
important. Data collection from injured patients, such as mechanism of
injury or mortality, are essential for creating databases that can be
used for research. In turn, that research can be used for planning
injury prevention programs that target the most common injuries in the
most efficient ways, e.g. teen drivers, children, specific occupations
etc. These injury prevention programs can be planned on trauma center,
organizational or government level[@ACS].

Trauma teams are multidisciplinary and operate in these centers. They
play a pivotal role in the treatment of the trauma patient, as they
provide the initial care in the critical stage of trauma. In Sweden,
trauma teams are lead by a team leader who is a surgeon, and include
practitioners from the specialities of intensive care, orthopedics,
nursing and support staff[@västra_G]. For the trauma team to be
mobilized, trauma code has to be activated, often by an emergency nurse.
The nurse uses information gathered by the first responders to assess if
the patient fits any of the criteria for trauma code. Most healthcare
facilities have established criteria or guidelines that trigger trauma
code activation. These criteria typically include mechanisms of injury,
physiological criteria, anatomical criteria and other specific
indicators of severe trauma. There are different levels of trauma code
with different criteria correlating with severity of injury, with level
1 mobilizing the most personnel to the trauma room. Once in the trauma
room, the team works systematically to manage the patients injuries.
They handle the most urgent problems first, such as airways and
breathing, with the aim to rapidly assess and stabilise the patient,
prioritise their injuries and arrange for site of definitive care.
[@Hedberg_2020]

## Mortality and morbidity (MoM) conferences

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology
[@SweTrau_2021] This logging of information is important work, as a
cornerstone of trauma quality improvement programs is multidisciplinary
MoM conferences. The MoM conference is a meeting where different
specialities who work with trauma care sit down and discuss deaths and
complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
mortality conferences on deceased patient cases, but only one has MoM
conferences to better their trauma care on patients who lived
[@SweTrau_2023]. The reason being that this expansion of MoM conferences
takes time and resources, which is why all hospitals don't implement
this.

@@

## Aim

The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

@@

Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To have your
information added to the trauma registry, a patient must be over 15
years of age, had ISS\>9 and/or had trauma code activation. They get
added to both the Karolinska Trauma registry as well as the national
trauma registry (SweTrau). The registry includes data on vital signs,
injuries, interventions and patient information. Another database, the
trauma care quality database includes information relevant for MoM
conferences. At Karolinska University Hospital, patients who die during
hospital stay gets their case sent to a MoM everytime. However, for the
rest of the patients, two specialized nurses determine whether they are.
Several audit filters are set up to flag for patients with possible
OFIs. Audit filters may include low SBP, high RR, EMS scene time greater
than 20 mins etc. Flagged patients are then reviewed by these two nurses
who decide if there is a possible OFI, in which case their case is sent
to a MoM conference for further review.

@@

The outcome is the ofi delayed treatment. An ofi is any failure in care,
e.g. clinical judgement error, any preventable or potentially
preventable death, missed diagnosis, technical error and delayed
treatment as identified and decided my the MoM conference. The outcomes
we'll be using is "long time until CT" and "long time until operation",
which will be referred to as delayed treatment from now on.

@@

The data comes from the trauma registry and trauma care quality
database,

@@

This process was the only time AI was used in the writing process. It
was used to make snippets of code for a specific purpose, revise and
optimize code and find solutions when written code was not functioning
properly.

@@

As we were running the logistical regression, further were excluded from
the study due to missing data. Among the 8,312 patients screened for
ofi, 7,378 were included in the study. 934 patients were excluded from
the study due to missing values in one or more variables. This left us
with 70 patients in the delayed treatment group and 7,308 patients in
the no delay to treatment group. The flowchart below illustrates the
exclusion and inclusion of patients.

@@

Below are the characteristics of the group used in the logistical
regression. Most of the patients were men, 68%, with that being a bit
higher in the delayed treatment group, 74%. The mean value for the ISS
score was 10.18, with that value being higher in the outcome group, at
16.81. The mean age of the patients was 45.18 years old, with it being
around the same in the delayed treatment group at 45.50. The mean number
of injuries among the delayed treatment patients was at 6.33, which is
higher than in the overall group at 4.76. The mean SBP was 137.33, with
a standard deviation of 25.69. Mean respiratory rate was 18.59 with it
being slightly higher in the outcome population at 20.39. Lastly, mean
GCS was 13.93, being lower among patients with delayed treatment at
12.56.

@@

In the following table, the results from the logistical regression
analysis is displayed.

The odds ratio is calculated both unadjusted and adjusted for other
variables, as well as a 95% confidence interval. The p value is also
shown.

The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate both have
a trend that patient with lower values seem to have a lower incidence of
delayed treatment, both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
OR 1.06, 95% CI 1.02-1.09) and adjusted (ISS, OR 1.03, 95% CI
1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
correlation, with higher GCS often having lower incidence of delayed
treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and adjusted (OR
0.93, 95% CI 0.87-1.00). Number of injuries had the correlation of
higher number of injuries having higher incidence of delayed treatment
unadjusted (OR 1.06, 95% CI 1.02-1.10), but that correlation was not
seen when adjusted for. The three remaining variables, age, gender and
SBP, showed no correlation with the observed outcome.

@@

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less awake patient is more
prone to receiving delayed treatment. Compared to previous a previous
study analysing the same data but with ofi as the outcome, this result
is similar to what they found using the same database [@Pat_factors]. As
the mentioned study suggests, this could be explained by that lower GCS
indicates severe trauma, as well as higher ISS. More injured patients
often entails a more complex management and require more interventions.
More interventions leave more room for delay to these interventions, why
they are more prone to delay in treatment. Respiratory rate also had the
correlation that lower respiratory rate correlated with lower rates of
delayed treatment among the patients. As the interval is \>10 to \<29 in
the triage guideline, both too low and too high respiratory rate is not
desirable[@EHCP2015]. This might have a similar explanation to ISS and
GCS, that patients with tachypnea might receive higher rates of delayed
treatment because they are more severely injured, therefore a higher
risk of missing a possible injury. A previous study also found this, but
the correlation dissapeared when adjusted for, which was hypothesised
being due to it's shared variability with the ISS
variable[@Pat_factors]. In this study, this was the case for number of
injuries, as it and ISS is often correlated as a more severely injured
patient will both score higher on ISS and number of injuries. Therefore,
we hypothesize that the effect of number of injuries become
insignificant when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, with many analysing mortality. Some
likeness exist with which patient level factors predict the outcome,
with higher mortality being seen in patients with lower GCS and higher
ISS [@Mort_sam][@Mort_hef]. Increased age was also a factor linked to
mortality, which was not seen to be linked to delayed treatment in this
study. An older patient is more likely to have co-morbidities that might
affect their mortality but not increase their likelihood of receiving
delayed treatment the same way ISS and GCS does[@Gioffre2018].

When looking at the patient demographic, it is clear that the majority
admitted to trauma care are men, with it being slightly higher in the
delay to treatment group. However, not high enough to be statistically
significant. The fact that gender did not show an increased odds ratio
should be seen as a positive sign. It suggests that among trauma
patients, there is no systematic delay based on this patient level
factor.

@@

In this study, one of the limitations was the amount of people having
the recorded outcome being quite small, leading us into having to limit
the amount of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories, from 1-4. However, many of the RTS variables had few to no
patients with the observed outcome. This led to separation in the
logistical regression in these variables, referring to when an outcome
variable separates a predictor variable completely, which occurs when
the predictor is only associated with one outcome value [@OARC]. In the
case of these variables, it was only RTS 3 and 4 who had patients with
the observed outcome, which left RTS 1 and 2 with confidence intervals
of 0.00 to infinity. This way would have been the best way to examine
these variables, but as that wasn't possible with this group of
patients, we resorted to making these variables continuous. We
considered expanding the outcome group to include more ofi, but opted
for the solution used in the study instead as we were solely interested
in this ofi for this study.

The problem mentioned above with too few patients in the outcome group
made this into a post hoc analysis. As we didn't know the amount of
people who received delayed treatment when we handled the scrambled
data, we ran the code on the real data when we had too many variables
for the dataset. Therefore, we had to change the variables after having
seen the real data, making it a post hoc analysis. This could in theory
open up for possible selection bias, as we might choose variables which
give us results we are looking for. However, all the variables included
in the final analysis were already included in the first one. Variables
were only removed, not added after having seen the real data. The best
way to have gone about this would have been to just look at the amount
of people who received delayed treatment before choosing the variables.
Then, choosing an amount of variables that would work with the amount of
patients. This would have avoided making it post hoc and changing the
variables after seeing the data.

On discussing possible biases, misclassification bias might also be
present in this study, though it is something that we can't control in
this study. As previously mentioned, the possible existence of an ofi is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

Another limitation with this study is that logistical regressions
assumes that the data is binomial, which the data used is not. The
analysis assumes a linear correlation between the dependent variable and
the independent variable, which is best analysed in a normal
distribution. However, Fischers test showed a significant difference
between the groups in 4 of the 7 variables, all of them being the ones
who showed an increased or decreased risk of delayed treatment, showing
that these results still are significant.

@@

@@

Discussing the generalisability of this study, we included the all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
ofi. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. This is in line with our aim, as it was to make it
applicable to the whole trauma patient population. However, as it is a
single center study, where it's result only illustrates the situation in
Stockholm, it is less applicable nationally or globally. This does not
make it only applicable to the population of Stockholm, but is something
to keep in mind.

@@

In future studies building upon this one, various methods to further
examine the correlation between delayed treatment and various variables,
alternatively using another ofi as the outcome variable. Examining
different variables such as highest hospital care lever, intubation
status or first emergency procedure performed might be interesting
variables, among others. To be able to do this, more patients in the
outcome group would be needed, either by expanding the register to
several hospitals, or expanding the outcome group by adding different
ofi. This however would make the study less specific for delayed
treatment. Further studies might also look into different ofi other than
delayed treatment, such as missed injury, faulty communication or
triage.

@@

In conclusion, based on this small population, the results suggests that
it is mostly more injured and less awake trauma patients that receive
delayed treatment. However, due to the limitations of this study, the
result should only be seen as something to keep in mind when treating
patients as further studies need to be conducted on this subject.

>>>>

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(knitr)
```

@@

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
<!-- Consider clarifying that delayed treatment may be one of the most 
common OFI-->
opportunities for improvement, e.g. delayed treatment. *Aim:* This
retrospective cohort study aims to identify patient level factors that
<!-- I would use "associated" rather than "linked"-->
are linked with receiving delayed treatment among trauma patients.
<!-- The first sentence is a bit complicated. Try to simplify, break 
into multiple sentences -->
*Material and Methods:* This study uses the database of Karolinska
university hospital, the equivalent of a level 1 trauma center, to run a
logistical regression linking the outcome, delayed treatment, to
variables, observed patient level factors. Patients treated for trauma
at the hospital between 2012 and 2022 were included, with ISS\>9 and/or
trauma code activation. Exclusion criterias were not being screened for
OFI and missing values in any of the variables, age, gender, ISS, SBP, RR,
GCS and number of injuries. *Results:* Higher ISS and RR was linked to
higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not when adjusted for. The
remaining variables showed no correlation. *Conclusion:* Our study
suggests that it is mostly more severely injured and less awake patients
that receive delayed treatment.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response [@gerdin_risk_2015], is and has long been
a major cause of death around the world [@WHO_2021]. 
<!-- The sentence below is long and overly complicated. Are you saying 
that trauma systems is an important part of healthcare and its implementation
has reduced mortality and morbidity?-->
Within modern
hospital care, the trauma system stands as an important part in
healthcare and is crucial to lowering mortality and morbidity in injured
patients through many means, such as hospital care, patient follow up
and prevention programs [@Choi_2021].

## Trauma Statistics
<!-- I suggest you name this section something else, like "Trauma globally" 
or something like that. Also, divide the second paragraph into two or 
three paragraphs-->
Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
groups, as blunt trauma patients tend to be more injured on arrival to
the hospital and as such, require more resources and are hospitalized
for a longer periods [@FITCH2019].

@@

## Trauma systems
<!-- Split long paragraphs -->
Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the
last three decades, the introduction of trauma systems has contributed
to lowering the incidence of preventable death. This is attributed to
the improvements in care for acute brain injuries and bleeding control.
The incidence of late death because of sepsis and multiple organ failure
has also been lowered, possibly a result of better and earlier
resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
the hospital, but their work outside of the hospital is just as
important. Data collection from injured patients, such as mechanism of
injury or mortality, are essential for creating databases that can be
used for research. In turn, that research can be used for planning
injury prevention programs that target the most common injuries in the
most efficient ways, e.g. teen drivers, children, specific occupations
etc. These injury prevention programs can be planned on trauma center,
organizational or government level[@ACS].

Trauma teams are multidisciplinary and operate in these centers. They
play a pivotal role in the treatment of the trauma patient, as they
provide the initial care in the critical stage of trauma. In Sweden,
trauma teams are lead by a team leader who is a surgeon, and include
practitioners from the specialities of intensive care, orthopedics,
nursing and support staff[@västra_G]. For the trauma team to be
mobilized, a trauma code has to be activated, often by an emergency nurse.
The nurse uses information gathered by the first responders to assess if
the patient fits any of the criteria for trauma code. Most healthcare
facilities have established criteria or guidelines that trigger trauma
code activation. These criteria typically include mechanisms of injury,
physiological criteria, anatomical criteria and other specific
indicators of severe trauma. There are different levels of trauma code
with different criteria correlating with severity of injury, with level
1 mobilizing the most personnel to the trauma room. Once in the trauma
room, the team works systematically to manage the patients injuries.
They handle the most urgent problems first, such as airways and
breathing, with the aim to rapidly assess and stabilise the patient,
prioritise their injuries and arrange for site of definitive care.
[@Hedberg_2020]

## Mortality and morbidity (MoM) conferences
<!-- Consider renaming this section as "Opportunities for improvement" -->
Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology
[@SweTrau_2021] This logging of information is important work, as a
cornerstone of trauma quality improvement programs is multidisciplinary
MoM conferences. The MoM conference is a meeting where different
specialities who work with trauma care sit down and discuss deaths and
complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
mortality conferences on deceased patient cases, but only one has MoM
conferences to better their trauma care on patients who lived
[@SweTrau_2023]. The reason being that this expansion of MoM conferences
takes time and resources, which is why all hospitals don't implement
this.

@@

## Aim
<!-- Make sure the aim here is exactly the same as in the abstract. Also,
I suggest that your aim is to determine the associations between different 
patient level factors and receiving delayed treatment -->
The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

@@

The Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To be included in
 the trauma registry, a patient must be over 15
years of age, had ISS\>9 and/or had trauma code activation. They get
added to both the Karolinska Trauma registry as well as the national
trauma registry (SweTrau). The registry includes data on vital signs,
injuries, interventions and patient information. Another database, the
trauma care quality database includes information relevant for MoM
conferences. At Karolinska University Hospital, patients who die during
hospital stay gets their case sent to a MoM everytime. However, for the
rest of the patients, two specialized nurses determine whether they are.
Several audit filters are set up to flag for patients with possible
OFIs. Audit filters may include low SBP, high RR, EMS scene time greater
than 20 mins etc. Flagged patients are then reviewed by these two nurses
who decide if there is a possible OFI, in which case their case is sent
to a MoM conference for further review.

@@

The outcome is the OFI delayed treatment. An OFI is any failure in care,
e.g. clinical judgement error, any preventable or potentially
preventable death, missed diagnosis, technical error and delayed
treatment as identified and decided my the MoM conference. The outcomes
we used were "long time until CT" and "long time until operation",
which will be referred to as delayed treatment from now on.

@@

The data comes from the trauma registry and trauma care quality
database.

@@

This process was the only time generative AI was used in the writing process. It
was used to make snippets of code for a specific purpose, revise and
optimize code and find solutions when written code was not functioning
properly.

@@

As we were running the logistic regression, additional patients were excluded from
the study due to missing data. Among the 8,312 patients screened for
ofi, 7,378 were included in the study. 934 patients were excluded from
the study due to missing values in one or more variables. This left us
with 70 patients in the delayed treatment group and 7,308 patients in
the no delay to treatment group. Figure 1 below illustrates the
exclusion and inclusion of patients.

@@

<!-- You need to refer to tables and figures by numbers in the text, 
i.e. Figure 1, Table 1 and so on -->
Table 1 shows the characteristics of the group used in the logistic
regression. Most of the patients were men, 68%, with that being a bit
higher in the delayed treatment group, 74%. The mean value for the ISS
score was 10.18, with that value being higher in the outcome group, at
16.81. The mean age of the patients was 45.18 years old, with it being
around the same in the delayed treatment group at 45.50. The mean number
of injuries among the delayed treatment patients was at 6.33, which is
higher than in the overall group at 4.76. The mean SBP was 137.33, with
a standard deviation of 25.69. Mean respiratory rate was 18.59 with it
being slightly higher in the outcome population at 20.39. Lastly, mean
GCS was 13.93, being lower among patients with delayed treatment at
12.56.

@@

In the following table<!-- See comments above-->, the results from the logistical regression
analysis is displayed.

The odds ratio is calculated both unadjusted and adjusted for other
variables<!-- What does "other" mean here?-->, as well as a 95% confidence interval. 
The p value is also shown.

<!-- Use "associated" instead of correlated or linked etc-->
The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate (RR) both have
a trend that patient with lower values seem to have a lower incidence of
delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR 1.03, 95% CI
1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
correlation, with higher GCS often having lower incidence of delayed
treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and adjusted (OR
0.93, 95% CI 0.87-1.00). Number of injuries had the correlation of
higher number of injuries having higher incidence of delayed treatment
unadjusted (OR 1.06, 95% CI 1.02-1.10), but that correlation was not
seen when adjusted for. The three remaining variables, age, gender and
SBP, showed no correlation with the observed outcome.

@@

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less awake patient is more
prone to receiving delayed treatment. Compared to a previous
study analysing the same data but with OFI as the outcome, this result
is similar to what they found using the same database [@Pat_factors]. As
the mentioned study suggests, this could be explained by that lower GCS as well as higher ISS
indicates severe trauma. More injured patients
often requires a more complex management and more interventions.
More interventions leave more room for delay to these interventions, why
they are more prone to delay in treatment. Respiratory rate also had the
correlation that lower respiratory rate correlated with lower rates of
delayed treatment among the patients. As the interval is \>10 to \<29 in
the triage guideline, both too low and too high respiratory rate is not
desirable[@EHCP2015]. This might have a similar explanation to ISS and
GCS, that patients with tachypnea might receive higher rates of delayed
treatment because they are more severely injured, therefore a higher
risk of missing a possible injury. The previous study also found this, but
the correlation dissapeared when adjusted for, which was hypothesised
being due to its shared variability with the ISS
variable[@Pat_factors]. In this study, this was the case for number of
injuries, as it and ISS is often correlated as a more severely injured
patient will both score higher on ISS and number of injuries. Therefore,
we hypothesize that the effect of number of injuries become
insignificant when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, while many analysing mortality. Some
similarities exist with which patient level factors predict the outcome,
with higher mortality being seen in patients with lower GCS and higher
ISS [@Mort_sam][@Mort_hef]<!-- Are these two references?-->. Increased age was also a factor linked to
mortality, which was not seen to be linked to delayed treatment in this
study. An older patient is more likely to have co-morbidities that might
affect their mortality but not increase their likelihood of receiving
delayed treatment the same way ISS and GCS does[@Gioffre2018].

When looking at the patient demographics, it is clear that the majority
admitted to trauma care are men, with it being slightly higher in the
delay to treatment group. However, not high enough to be statistically
significant. The fact that gender did not show an increased odds ratio
should be seen as a positive sign. It suggests that among trauma
patients, there is no systematic delay based on this patient level
factor.

@@

In this study, one of the limitations was the number of people having
the recorded outcome being quite small, leading us into having to limit
the number of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories, from 1-4. However, many of the RTS variables had few to no
patients with the observed outcome. This led to separation in the
logistic regression in these variables, referring to when an outcome
variable separates a predictor variable completely, which occurs when
the predictor is only associated with one outcome value [@OARC]. In the
case of these variables, it was only RTS 3 and 4 who had patients with
the observed outcome, which left RTS 1 and 2 with confidence intervals
of 0.00 to infinity. This way would have been the best way to examine
these variables, but as that was not possible with this group of
patients, we resorted to making these variables continuous. We
considered expanding the outcome group to include more OFI, but opted
for the solution used in the study instead as we were solely interested
in this OFI for this study.

The problem mentioned above with too few patients in the outcome group
made this into a post hoc analysis. As we did not know the number of
people who received delayed treatment when we handled the scrambled
data, we ran the code on the real data when we had too many variables
for the dataset. Therefore, we had to change the variables after having
seen the real data, making it a post hoc analysis. This could in theory
open up for possible selection bias, as we might choose variables which
give us results we are looking for. However, all the variables included
in the final analysis were already included in the first one. Variables
were only removed, not added after having seen the real data. The best
way to have gone about this would have been to just look at the amount
of people who received delayed treatment before choosing the variables.
Then, choosing an amount of variables that would work with the amount of
patients. This would have avoided making it post hoc and changing the
variables after seeing the data.

On discussing possible biases, misclassification bias might also be
<!-- Do not use contractions, i.e. can't, doesn't etc -->
present in this study, though it is something that we cannot control in
this study. As previously mentioned, the possible existence of an OFI is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

<!-- Not so sure about this section. What do you mean that the logistic
regression assumes that the data is binomial? Do you mean that it requires 
a binary outcome? I think you discuss potential limitations well enough in 
the previous paragraphs, so you don't really need this -->
Another limitation with this study is that logistic regressions
assumes that the data is binomial, which the data used is not. The
analysis assumes a linear correlation between the dependent variable and
the independent variable, which is best analysed in a normal
distribution. However, Fischers test showed a significant difference
between the groups in 4 of the 7 variables, all of them being the ones
who showed an increased or decreased risk of delayed treatment, showing
that these results still are significant.

@@

<!-- Revise this section now after our meeting -->

@@

Discussing the generalisability of this study, we included the all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
ofi. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. This is in line with our aim, as it was to make it
applicable to the whole trauma patient population. However, as it is a
single center study, where its result only illustrates the situation in
Stockholm, it is less applicable nationally or globally. This does not
make it only applicable to the population of Stockholm, but is something
to keep in mind.

<!-- Think about what components of your study is generalisible and to 
where. I think that your results are generalisible to  a large extent, in 
the sense that severly injured patients are at higher risk of delayed 
treatment, also in other contexts -->

@@

In future studies building upon this one, various methods to further
examine the correlation between delayed treatment and various variables,
alternatively using another OFI as the outcome variable. Examining
different variables such as highest hospital care lever, intubation
status or first emergency procedure performed might be interesting
variables, among others. To be able to do this, more patients in the
outcome group would be needed, either by expanding the register to
several hospitals, or expanding the outcome group by adding different
ofi. This however would make the study less specific for delayed
treatment. Further studies might also look into different ofi other than
delayed treatment, such as missed injury, faulty communication or
triage.

@@

In conclusion, based on this small population, the results suggests that
<!-- I would say "with affected consciousness" instead of "less awake" -->
it is mostly more injured and less awake trauma patients that receive
delayed treatment. However, due to the limitations of this study, the
result should only be seen as something to keep in mind when treating
patients as further studies need to be conducted on this subject.
